---
title: "Staph-Survival-Paper"
author: "Maria Steadmon"
date: "1/22/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Statistics and analysis for the manuscript titled "Comparing the survival and persistence of *Staphylococcus aureus* in environmental waters of different turbidity"

**Libraries required**

```{r}
library(plyr)
library(ggplot2)
library(MuMIn)
library(MASS)
library(car)
library(effects)
```

**Data required**

```{r}

setwd("/Users/msteadd/Documents/") #change to your own working directory of where the data is stored
dat=read.csv("Staph_Survival_Data_Raw.csv", na.strings = ".", header=TRUE) 
head(dat)
dat$Site=as.factor(dat$Site)
```

**Environmental (initial) concentrations of *S. aureus***

```{r}
#Subset
initial_data <- subset(dat,Time_day %in% c("0"))
```

**Metadata for initial concentrations of *S. aureus*** \newline

**Freshwater**

```{r}
#initial_data
fresh_initial_data___ <- subset(initial_data,Salinity %in% c("Fresh"))

initial_data_averages_fresh <- ddply(fresh_initial_data___, .(Turbid_Level), summarize, avg=mean(Staph_CFU_100mL), Median = median(Staph_CFU_100mL), 
                              min=min(Staph_CFU_100mL),
                              max=max(Staph_CFU_100mL),
                              IQR = IQR(Staph_CFU_100mL),
                            SE=sd(Staph_CFU_100mL)/sqrt(length(Staph_CFU_100mL)),
                            SD = sd(Staph_CFU_100mL))
initial_data_averages_fresh
```

**Brackish water**

```{r}
brackish_initial_data___ <- subset(initial_data,Salinity %in% c("Brackish"))

initial_data_averages_brackish <- ddply(brackish_initial_data___, .(Turbid_Level), summarize, avg=mean(Staph_CFU_100mL), Median = median(Staph_CFU_100mL), 
                              min=min(Staph_CFU_100mL),
                              max=max(Staph_CFU_100mL),
                              IQR = IQR(Staph_CFU_100mL),
                            SE=sd(Staph_CFU_100mL)/sqrt(length(Staph_CFU_100mL)),
                            SD = sd(Staph_CFU_100mL))
initial_data_averages_brackish
```

**Means test for comparing *S. aureus* concentrations**

```{r}
initial_data$Turbid_Level=as.factor(initial_data$Turbid_Level)
initial_data$Salinity=as.factor(initial_data$Salinity)
initial_data$Sal_Turb=as.factor(initial_data$Sal_Turb)


kruskal.test(initial_data$Staph_CFU_100mL~initial_data$Sal_Turb)
pairwise.wilcox.test(initial_data$Staph_CFU_100mL,initial_data$Sal_Turb, p.adjust.method = "BH")

```

**Figure 1**

```{r}

my_y_title_staph_<-expression(paste(italic("S. aureus"), " CFU/100 mL"))

grayscale=c("black","#999999")

  Figure1=ggplot(initial_data, aes(x=Turbid_Level,y=Staph_CFU_100mL,color=Salinity))+
  geom_boxplot()+theme_bw()+
  #geom_point()+
  theme(panel.grid=element_blank(), legend.position="top", text=element_text(size=20))+ 
    xlab("Turbidity Level")+
  ylab(my_y_title_staph_)+#scale_fill_manual(name="Turbidity Level",values=cbPalette)+
 # scale_color_manual(values=cbPalette)+
      coord_cartesian(ylim=c(0, 13000))+
    scale_fill_grey()+
    scale_color_manual(name="Salinity",values=grayscale)#+
  #facet_wrap(~ salinity, scales="free")#, nrow=1, ncol=2)
  Figure1
```

**Models of environmental concentrations of *S. aureus***

```{r}
#Full model
model_initial_data=lm(log(Staph_CFU_100mL+1)~ 
                      Salinity+Turbid_Level+
                      (1|Bottle), data=initial_data)
#No salinity
model_initial_data_sal=lm(log(Staph_CFU_100mL+1)~
                    Turbid_Level+(1|Bottle), data=initial_data)

#No turbidity
model_initial_data_turbid=lm(log(Staph_CFU_100mL+1)~ 
                      Salinity+(1|Bottle), data=initial_data)

#F-Test
Anova(model_initial_data)

#Comparing all models to one another
staph_survival_models_initial=list(model_initial_data,
                          model_initial_data_sal,
                           model_initial_data_turbid)
model.sel(staph_survival_models_initial)
```

***S. aureus*** **survival calculations and models**

**Model for fresh water, low turbidity**

```{r}
#log base 10
Fresh_data <- subset(dat,Salinity %in% c("Fresh"))
low_turbidity_Fresh_data <- subset(Fresh_data,Turbid_Level %in% c("Low"))


model_for_T90_low_turbidity_Fresh <- lm((log(Staph_CFU_100mL+1)) ~ Time_day + (1 | Bottle), data = low_turbidity_Fresh_data)
summary(model_for_T90_low_turbidity_Fresh)
plot(allEffects(model_for_T90_low_turbidity_Fresh))

coef(model_for_T90_low_turbidity_Fresh)
Anova(model_for_T90_low_turbidity_Fresh)
```

**Model for fresh water, high turbidity**

```{r}
#log base 10
#Fresh_data <- subset(dat,Salinity %in% c("Fresh"))
high_turbidity_Fresh_data <- subset(Fresh_data,Turbid_Level %in% c("High"))


model_for_T90_high_turbidity_Fresh <- lm((log(Staph_CFU_100mL+1)) ~ Time_day + (1 | Bottle), data = high_turbidity_Fresh_data)
summary(model_for_T90_high_turbidity_Fresh)
plot(allEffects(model_for_T90_high_turbidity_Fresh))

coef(model_for_T90_high_turbidity_Fresh)
Anova(model_for_T90_high_turbidity_Fresh)
```

**Model for brackish water, low turbidity**

```{r}
#log base 10
brackish_data <- subset(dat,Salinity %in% c("Brackish"))
low_turbidity_brackish_data <- subset(brackish_data,Turbid_Level %in% c("Low"))


model_for_T90_low_turbidity_brackish_lm <- lm((log(Staph_CFU_100mL+1)) ~ Time_day + (1 | Bottle), data = low_turbidity_brackish_data)
summary(model_for_T90_low_turbidity_brackish_lm)
plot(allEffects(model_for_T90_low_turbidity_brackish_lm))

coef(model_for_T90_low_turbidity_brackish_lm)
Anova(model_for_T90_low_turbidity_brackish_lm)
```

**Model for brackish water, intermediate turbidity**

```{r}
int_turbidity_brackish_data <- subset(brackish_data,Turbid_Level %in% c("Intermediate"))
#high turbidity brackish data
model_for_T90_int_turbidity_brackish_lm <- lm((log(Staph_CFU_100mL+1)) ~ Time_day + (1 | Bottle), data = int_turbidity_brackish_data)
summary(model_for_T90_int_turbidity_brackish_lm)
plot(allEffects(model_for_T90_int_turbidity_brackish_lm))

coef(model_for_T90_int_turbidity_brackish_lm)
Anova(model_for_T90_int_turbidity_brackish_lm)
```

**Model for brackish water, high turbidity**

```{r}

high_turbidity_brackish_data <- subset(brackish_data,Turbid_Level %in% c("High"))

#high turbidity brackish data
model_for_T90_high_turbidity_brackish_lm <- lm((log(Staph_CFU_100mL+1)) ~ Time_day + (1 | Bottle), data = high_turbidity_brackish_data)
summary(model_for_T90_high_turbidity_brackish_lm)
plot(allEffects(model_for_T90_high_turbidity_brackish_lm))

coef(model_for_T90_high_turbidity_brackish_lm)
Anova(model_for_T90_high_turbidity_brackish_lm)
```

**Figure 2 code**

```{r}
#data averaged over 6 replicates for figure 2
setwd("/Users/msteadd/Documents/Staph_Survival")
dat1=read.csv("staph_survival_data.csv", na.strings = ".", header=TRUE)
head(dat1)
dat1$site=as.factor(dat1$site)
```

```{r}
my_y_title_staph_sand<-expression(paste("Log ", italic("S. aureus"), " CFU/100 mL"))

Figure2=ggplot(dat1, aes(x=time,y=log(staph),
                        fill=turbid_level,color=turbid_level))+ 
  geom_point(aes(shape=turbid_level))+
  geom_smooth(method="lm",se=TRUE,aes(fill=turbid_level))+
  xlab("Time (Days)")+
  ylab(my_y_title_staph_sand)+
  scale_fill_grey(name="Turbidity Level")+
  scale_color_grey(name="Turbidity Level")+
  theme_bw()+
        theme(panel.grid.major = element_blank(),legend.position = "top",
          panel.grid.minor = element_blank(),        
          legend.text=element_text(size=25),
          axis.line  = element_line(colour = "black")) +
  facet_wrap(~salinity, nrow=2, ncol=2)
Figure2
```
